A Handbook for Tenured Track Faculty in Computer Science in the US

### 2024-10-16 demystify.tex
updates

    % Title
    \node[align=center] at ($(current page.center)+(0,-7)$)
    {
    {\fontsize{24}{24} \selectfont {{Demystifying the Computer Science}}} \\[0.15in]
    {\fontsize{24}{24} \selectfont {{PhD Admission in the US}}} \\[1in]    
    %{\fontsize{18}{18} \selectfont {{A Handbook for International Students}}} \\[1.in]

@@

% though \emph{very} top schools could be very selective, e.g., see the \href{https://da-data.blogspot.com/2015/03/reflecting-on-cs-graduate-admissions.html}{admission process} at CMU
So I want to share some details about the admission process and advice for those who are interested in applying for a \textbf{PhD in Computer Science in the US}.
Originally, this document was intended for \emph{international students}, but I have expanded it to include information that might also be useful for \emph{US domestic students}.
Moreover, while this primarily aims at students interested in CS, it might be relevant to students from various STEM (Science, Technologies, Engineering, and Mathematics) disciplines.
Furthermore, although many examples are specifics for schools that I and other contributors of this document know about, the information should be generalizable to other good R1\footnote{An \href{https://en.wikipedia.org/wiki/List_of_research_universities_in_the_United_States}{R1 institution} in the US is a research-intensive university with a high level of research activity across various disciplines. Currently, 146 (out of 4000) US universities are classified as R1.} institutions in the US.

@@

I wish you the best of luck.

\begin{mybox}
This document is open source and updated regularly. Its latest version is available at

@@

First, applying to a good US university \emph{should not} be any harder than at schools in other countries. In fact, it might even be more flexible since CS PhD in the US \emph{do not} require having an MS or a research topic, proposal, or adviser in advance (\autoref{sec:non-us-differences} compares CS PhD study in the US to other countries). If you believe you have a chance in other countries, e.g., South Korea, Singapore, Germany, UK, Japan, and Australia, then you will surely have a chance in the US as well.

Second, PhD students in CS \emph{do not} need to worry about funding, especially at good R1
universities in the US. If you are admitted, you will almost certainly
receive \emph{full funding} (\autoref{sec:funding}) to support your study. This includes tuition,
health insurance, and stipend (i.e., you get paid for your study). Moreover, depending on the university,
you may receive additional benefits such as summer pay, laptops, and traveling to conferences and workshops.

@@

Students from top schools with strong research cultures and experience might already know most of the information given here.
My goal is thus to level the playing field by providing information that is not readily available to less privileged students. 
I hope to encourage more students with such backgrounds to apply to CS PhD programs in the US.

@@

Typically, it takes 5--7 years for CS PhD in the US.  This can be longer than CS PhDs in other countries (\autoref{sec:msrequirement}), which might require MS first. Within these 5--7 years, CS PhD students often take a ``leave of absence'' for 1--2 semesters to do internships at companies and research labs.

@@

% \begin{table}[h!]
% \centering
% \begin{tabular}{l|l}
%     \toprule
% \textbf{Month} & \textbf{Task} \\
% \midrule
% January-March & Research programs and prepare for tests \\
% April-June & Take standardized tests (GRE, TOEFL, etc.) \\
% July-September & Prepare application materials (SOP, LORs, etc.) \\
% October-December & Submit applications \\
% January-March & Interviews and follow-ups \\
% April & Admission decisions \\
% \bottomrule
% \end{tabular}
% \caption{Application Timeline}
% \end{table}
% \section{Why the US and not other countries?}
% TODO

@@

\section*{Additional Resources}
\begin{itemize}
  \item \href{https://parentheticallyspeaking.org/articles/us-cs-phd-faq/}{Getting a Computer Science PhD in the USA} by Shriram Krishnamurthi
  \item  \href{https://pg.ucsd.edu/PhD-application-tips.htm}{Blog post on PhD Admission} by Philip Guo
  \item   \href{https://talkingtorobots.com/FAQ.html}{PhD Admission post} by Yonatan Bisk (customized for CMU admission)
\end{itemize}

@@

\subsection{Why CS depts do not waive the application fee?}

>>>>

    % Title
    \node[align=center] at ($(current page.center)+(0,-7)$)
    {
    {\fontsize{24}{24} \selectfont {{Demystifying the Computer Science}}} \\[0.15in]
    {\fontsize{24}{24} \selectfont {{PhD Admission in the US}}} \\[0.2in]    
    {\fontsize{18}{18} \selectfont {{A Handbook for International Students}}} \\[1.in]

@@

% though \emph{very} top schools could be very selective, e.g., see the \href{https://da-data.blogspot.com/2015/03/reflecting-on-cs-graduate-admissions.html}{admission process} at CMU
So I want to share some details about the admission process and advice for those who are interested in applying for a \textbf{PhD in Computer Science in the US}.
Originally, this document was intended for international students, but I have expanded it to include information that might also be useful for \emph{US domestic students}.
Moreover, while this is primarily intended for students interested in CS, it might be relevant to students from various STEM (Science, Technologies, Engineering, and Mathematics) disciplines.
Furthermore, although many examples are specifics for schools that I and other contributors of this document know about, the information should be generalizable to other R1\footnote{An \href{https://en.wikipedia.org/wiki/List_of_research_universities_in_the_United_States}{R1 institution} in the US is a research-intensive university with a high level of research activity across various disciplines. Currently, 146 (out of 4000) US universities are classified as R1.} institutions in the US.

@@

I wish you the best of luck. Happy school hunting!

\begin{mybox}
This document will be updated regularly to reflect the latest information and updates in the admission process. Its latest version is available at

@@

First, applying to a good US university \emph{should not} be any harder than at schools in other countries. In fact, it might even be more flexible since CS PhD in the US \emph{do not} require having an MS or a research topic/proposal/adviser in advance. If you believe you have a chance in other countries, e.g., South Korea, Singapore, Germany, UK, Japan, and Australia, then you will surely have a chance in the US as well. \autoref{sec:non-us-differences} compares CS PhD study in the US to other countries.

Second, PhD students in CS \emph{do not} need to worry about funding, especially at good R1
universities in the US. If you are admitted, you will almost certainly
receive \emph{full funding} (\autoref{sec:funding}) to support your study, including tuition,
health insurance, and stipend (i.e., you get paid for your study). Moreover, depending on the university,
you may even receive additional benefits such as summer pay, laptops, and traveling to conferences and workshops.

@@

Students from top schools with strong research cultures and experiences might already know most of the information given here.
My goal is thus to level the playing field by providing information that is not readily available to less privileged students. 
I hope to encourage more students with such backgrounds to apply to CS PhD programs in the US.

@@

Typically, it takes 5--7 years for CS PhD in the US.  This can be longer than CS PhDs in other countries, which might require MS first (recall from \autoref{sec:msrequirement} that CS PhD programs in the US do not require MS and you can get MS along the way to PhD). Within these 5--7 years, CS PhD students often take a ``leave of absence'' for 1--2 semesters to do internships at companies and research labs.

@@

% \begin{table}[h!]
% \centering
% \begin{tabular}{l|l}
%     \toprule
% \textbf{Month} & \textbf{Task} \\
% \midrule
% January-March & Research programs and prepare for tests \\
% April-June & Take standardized tests (GRE, TOEFL, etc.) \\
% July-September & Prepare application materials (SOP, LORs, etc.) \\
% October-December & Submit applications \\
% January-March & Interviews and follow-ups \\
% April & Admission decisions \\
% \bottomrule
% \end{tabular}
% \caption{Application Timeline}
% \end{table}
% \section{Why the US and not other countries?}
% TODO
\subsubsection{Why CS depts do not waive the application fee?}

@@

@@

### 2024-10-16 faculty.tex
updates

While teaching takes about 40\% of your time, it is not as nearly as important as research for tenure.  Moreover, most new faculty do not prioritize teaching, e.g., compared to writing grants or recruiting students for research.  However, teaching is still important for your tenure case, and being a good teacher can be very rewarding. Innovative integration of research and teaching can also help you get good students and even NSF grants.

\section{Teaching Load}
New TT faculty often have a lighter teaching load in their first several years or until they get tenure.  This is to give you time to establish your research program.

At most large R1 universities the teaching load for TT faculty is 1+1 (1 course / semester). While some univs might give you a lighter load in your first year, most univs expect you to teach 1+1 starting your second year. In many cases, 1+1 is only for the first 2 or 3 years and after that it increases to 1+2 (or 2+1). It is rare that a TT in CS at R1 universities would have to teach 2+2 during the tenure track.

In some cases your contract might list 1+1 for the first X year and 1+2 after until tenure, but the dept unofficial gives TT faculty 1+1 until tenure. Some other ways to reduce teaching load include \emph{buying out} of teaching with a grant, leading a very large grant, and teaching many large classes.

\begin{commentbox}[Vu]
    At UNL, TT faculty is 1+1 for the first 3 years, and 1+2 after that.  At GMU, it is 1+1 for the first 3 years, but most TT faculty would continue this teaching load until they get tenure. In addition, GMU grants all TT faculty a semester off from teaching before tenure. Some faculty use this to work on their NSF CAREER proposal while others use it to prepare their tenure application. 
\end{commentbox}

\section{Courses}

In most cases you will teach both undergrad and grad courses. New faculty often create graduate seminars and special topic courses to recruit students for their research group. Undergrad courses typically include higher level courses in your area of expertise (e.g., 400 level course on Software Testing). It is rare that you have to teach 100 level courses (e.g., Intro to Programming) because they are very large and thus time-consuming.

Most depts. would allow faculty to pick their favorite courses to teach. In some cases, they let  you develop your own grad course (e.g., Software Verification) but ask you to teach an existing core undergraduate courses (e.g., Theory, Operating Systems).
It is likely that you get to repeat the same courses. This significant reduces the amount of time you need to prepare for the course.  However, you should not teach the same course every semester, and you will want to alternate between grad and undergrad courses.

@@

    I feel that UNL is a bit more flexible in creating new classes compared to GMU.  However, I spent much less time prepare for courses at GMU.
  \end{commentbox}

@@

Teaching evaluations are a critical component of your annual review and P\&T. It provides an analysis on how students and peers perceive your teaching. 
You pay attention to evaluations, and make changes to your teaching based on the feedback, and show that you continuously strive to improve your teaching.

\paragraph{Student Evaluation}
At the end of each semester, students are asked to evaluate the instructor. The evaluation often includes both quantitative (e.g., ratings on a scale of 1-5) and qualitative (e.g., comments) feedback.  In addition to default questions, you can add your own questions. The results are shared anonymously with the instructor to help them improve their teaching.

One main issue with teaching evaluation is that not many students respond, despite multiple reminders and urging from the instructor.  Moreover, the few that respond are usually the students who are unhappy and thus will give highly negative evaluation. This is the reason why many instructors do not like teaching evaluations, especially when it is used to evaluate their performance.

\begin{commentbox}[Vu]
  A strategy that I found effective is that the whole class gets a small bonus point if everyone completes the evaluation. The bonus is calculated as the percentage of responses multiplied by a factor (X). So, if X is 0.5 and everyone participates, they get 0.5 points each. If only half respond, it's 0.25 points each. It might seem small, but it can make a difference for borderline grades. Since I don't curve grades, this is their only chance to boost their scores. Many students have benefited from this in my previous classes, which really grabs their attention. Plus, since evaluations happen before finals and grades aren't finalized, everyone's motivated to participate—they don't want to miss out. Students not only do it themselves but also remind their friends. 
  \\

  It's been working well for me: Spring '24 I had a 90\% response rate (25 out of 28 students), and Fall '23 was a perfect 100\% (16 out of 16 students). Before, my response rate was about 50\%, which is already higher than the average.
  \end{commentbox}

@@

In addition to student feedback, peer evaluations can offer valuable insights on your teaching. At many places, peer evaluations are required for P\&T.
Request evaluations from experienced colleagues who can observe your class and provide constructive criticism. Like with student evaluation, incorporating these peer feedback into your teaching and describe your changes and improvements.  Peer evaluation letters can also help you get Teaching Awards and other recognitions.

>>>>

While teaching takes about 40\% of your time, it is not as nearly as important as research for tenure.  Moreover, most TT faculty do not prioritize teaching compared to writing grants or working with students on papers.  However, teaching is still important for your tenure case, and being a good teacher can be very rewarding. Innovative integration of research and teaching can also help you get good students and even NSF grants.

\section{Courses Taught}\label{sec:teaching-load}

% At most large R1 universities the teaching load for TT faculty is 1+1 (1 course / semester). While some univs might give you a lighter load in your first year, most univs expect you to teach 1+1 starting your second year. In many cases, 1+1 is only for the first 2 or 3 years and after that it increases to 1+2 (or 2+1). It is rare that a TT in CS at R1 universities would have to teach 2+2 during the tenure track.
% In some cases your contract might list 1+1 for the first X year and 1+2 after until tenure, but the dept unofficial gives TT faculty 1+1 until tenure. Some other ways to reduce teaching load include \emph{buying out} of teaching with a grant, leading a very large grant, and teaching many large classes.

Most places want to see that you have taught both undergrad and grad courses by the time you go up for tenure.  A good dept. would allow and encourage faculty to pick their favorite courses to teach. In some cases, they let you develop your own grad course (e.g., Software Verification) but ask you to teach an existing core undergraduate one (e.g., Theory, Operating Systems). A good dept. also "protects" you from teaching intro courses (e.g., Intro to Programming), which are usually very large and the most time-consuming to teach.

It is likely that you get to repeat the same courses, which significant reduces the amount of time you need to prepare for the course.  However, you should not teach the same course every semester, and you will want to alternate between grad and undergrad courses.

@@

 THus UNL is a bit more flexible in creating new classes compared to GMU.  However, I spent much less time prepare for courses at GMU.
\end{commentbox}

\paragraph{Teaching load} New TT faculty often have a lighter teaching load in their first several years or until they get tenure.  This is to give you time to establish your research program.
\begin{commentbox}[Vu]
  At UNL, TT faculty is 1+1 for the first 3 years, and 1+2 after that.  At GMU, it is 1+1 for the first 3 years, but most TT faculty would continue this teaching load until they get tenure. In addition, GMU grants all TT faculty a semester off from teaching before tenure. Some faculty use this to work on their NSF CAREER proposal while others use it to prepare their tenure application. 
\end{commentbox}

@@

Teaching evaluations are a critical component of P\&T, and you will include these in your dossier when you go up for tenure. 
It provides a quantitative measure on how students and peers perceive your teaching.
You should pay attention to evaluations, and make changes to your teaching based on the feedback, and show that you continuously strive to improve your teaching.

\paragraph{Student Evaluation}
At the end of each semester, students are asked to evaluate the instructor. The evaluation often includes both quantitative (e.g., ratings on a scale of 1-5) and qualitative (e.g., comments) feedback.  In addition to default questions, you can add your own questions. The results are shared anonymously with the instructor to help them improve their teaching. 
You will want to get good evaluations, ``above'' the average of the dept or college (which might be challenging because you are compared to all faculty, including teaching faculty who are very good at teaching).

One issue with teaching evaluation is that not many students respond, and the few that do are usually the students who are unhappy and thus will give highly negative evaluation. This is the reason why many instructors do not like teaching evaluations, especially when it is used to evaluate their performance.

\begin{commentbox}[Vu]
  In recent years I encourage student to do teaching evaluation by giving the whole class a small bonus point if everyone completes the evaluation. The bonus is calculated as the percentage of responses multiplied by a factor (X). So, if X is 0.5 and everyone participates, they get 0.5 points each. If only half respond, it's 0.25 points each. It might seem small, but it can make a difference for borderline grades. Since I don't curve grades, this is their only chance to boost their scores. Moreover, since evaluations happen before finals and grades aren't finalized, everyone's motivated to participate as they don't want to miss out. Students not only do it themselves but also remind their friends. 
  \\

  This works well for me: Spring '24 I had a 90\% response rate (25 out of 28 students), and Fall '23 was a perfect 100\% (16 out of 16 students). Before, my response rate was about 50\%, which is already higher than the average.
  \end{commentbox}

@@

In addition to student feedback, there is peer evaluation where a more experienced faculty member observes your class and provides feedback.  This is usually required for P\&T (e.g., both GMU and UNL require it).  Like with student evaluation, incorporating feedback from peer evaluation into your teaching and describe your changes and improvements.  Peer evaluation letters can also help you get Teaching Awards and other recognitions.
